---
title: "Exercise3.2"
author: "M Mendez"
date: "2025-09-25"
output: html_document
---
Show the distributions of values visually. Explain, in comments, what those distributions mean statistically. Be thorough.
What do the distributions mean statistically? 
I have taken the mean, minimum, and maximum of each quantitative value . These graphs allow the user to see the range of Total Sales from each OrderId. I have added a line on each to display the mean, min, and max. 
```{r}
library(tidyverse)
library(dplyr)
sales <- read.csv('sales_data_sample.csv')

#TotalSales 
unitmean <- mean(sales$TotalSales)
unitmin <- min(sales$TotalSales)
unitmax <- max(sales$TotalSales)

# hist(sales$TotalSales, 
#      xlab = "Mean Sales",
#      main = paste("Sales"))
# abline( v = unitmean, col = 'red', lwd =2)
# 
# print(unitmean)
# 
# hist(sales$TotalSales, 
#      xlab = "Minimum Unit Sales",
#      main = paste("Minimum Sales"))
# abline( v = unitmin, col = 'red', lwd = 2)
# 
# print(unitmin)
# 
# hist(sales$TotalSales, 
#      xlab = "Maximum Unit Sales",
#      main = paste("Maximum Sales"))
# abline( v = unitmax, col = 'red', lwd = 2)
# 
# print(unitmax)

ggplot(sales, aes(x = TotalSales)) +
  ggtitle('Total Sales') +
  labs(subtitle = 'Mean, Min, and Max') +
  geom_histogram(fill = 'gray') +
  geom_vline(linetype = 'dashed', color = 'blue', xintercept = mean(sales$TotalSales)) +
  geom_vline(linetype = 'dashed', color = 'orange', xintercept = min(sales$TotalSales)) +
  geom_vline(linetype = 'dashed', color = 'green', xintercept = max(sales$TotalSales)) +
  labs(caption = "Blue line is the mean, Orange is min, and Green is the max.")

print(unitmin)
print(unitmax)
print(unitmean)

#Quantity
unitmin1 <- min(sales$Quantity)
unitmax1 <- max(sales$Quantity)
unitmean1 <- mean(sales$Quantity)

# hist(sales$Quantity, 
#      xlab = " Mean Qty",
#      main = paste("Qty"))
# abline( v = unitmean1, col = 'red', lwd = 2)
# 
# print(unitmean1)
# 
# hist(sales$Quantity, 
#      xlab = "Minimum Qty",
#      main = paste("Minimum Qty"))
# abline( v = unitmin1, col = 'red', lwd = 2)
# 
# print(unitmin1)
# 
# hist(sales$Quantity, 
#      xlab = "Maximum Qty",
#      main = paste("Maximum Qty"))
# abline( v = unitmax1, col = 'red', lwd = 2)

# print(unitmax1)

ggplot(sales, aes(x = Quantity)) +
  ggtitle('Quantity') +
  labs(subtitle = 'Mean, Min, and Max') +
  geom_histogram(fill = 'gray') +
  geom_vline(linetype = 'dashed', color = 'blue', xintercept = mean(sales$Quantity)) +
  geom_vline(linetype = 'dashed', color = 'orange', xintercept = min(sales$Quantity)) +
  geom_vline(linetype = 'dashed', color = 'green', xintercept = max(sales$Quantity)) +
  labs(caption = "Blue line is the mean, Orange is min, and Green is the max.")

print(unitmin1)
print(unitmax1)
print(unitmean1)

#UnitPrice
unitmin2 <- min(sales$UnitPrice)
unitmax2 <- max(sales$UnitPrice)
unitmean2 <- mean(sales$UnitPrice)

ggplot(sales, aes(x = UnitPrice)) +
  ggtitle('Unit Price') +
  labs(subtitle = 'Mean, Min, and Max') +
  geom_histogram(fill = 'gray') +
  geom_vline(linetype = 'dashed', color = 'blue', xintercept = mean(sales$UnitPrice)) +
  geom_vline(linetype = 'dashed', color = 'orange', xintercept = min(sales$UnitPrice)) +
  geom_vline(linetype = 'dashed', color = 'green', xintercept = max(sales$UnitPrice)) +
  labs(caption = "Blue line is the mean, Orange is min, and Green is the max.")

print(unitmin2)
print(unitmax2)
print(unitmean2)
# hist(sales$UnitPrice, 
#      xlab = "Mean Unit Price",
#      main = paste("Unit Price"))
# abline( v = unitmean2, col = 'red', lwd = 2) +
# 
# 
# print(unitmean2)
# 
# hist(sales$UnitPrice, 
#      xlab = "Minimum Price",
#      main = paste("Minimum Price"))
# abline( v = unitmin1, col = 'red', lwd = 2)
# 
# print(unitmin1)
# 
# hist(sales$UnitPrice, 
#      xlab = "Maximum Price",
#      main = paste("Maximum Price"))
# abline( v = unitmax1, col = 'red', lwd = 2)
# 
# print(unitmax1)

```
Show the values of central tendency, both numerically and visually. Interpret those values in your own words in Markdown.
Mode of Product shows the frequency of the product 
Mean of Unit Prices provides insight to the mean price by the units provided
Median of Total Sales provies insight to show the total sales are a little less than $500
```{r}

prodsales <- sales %>%
  group_by(Product) %>%
  summarise(mean(TotalSales))

#assistance from https://www.youtube.com/watch?v=y5yFOVbxTIg 
#Mode
MProd <- table(sales$Product)
Mqty <- table(sales$Quantity)
MTotal <- table(sales$TotalSales)
MUnit <- table(sales$UnitPrice)


barplot(height=MProd, names = prodsales$Product,
        xlab = "Product (Mode)", 
        ylab = 'Sales',
     main = paste("Mode of Product"))

print(MProd)

barplot(height=Mqty,
        xlab = "Quantity (Mode)", 
        ylab = 'Sales',
     main = paste("Mode of Quantity"))

print(Mqty)

barplot(height=MTotal,
        xlab = "Total (Mode)", 
        ylab = 'Sales',
     main = paste("Mode of Total"))

print(MTotal)

barplot(height=MUnit,
        xlab = "Unit Price (Mode)", 
        ylab = 'Sales',
     main = paste("Mode of Unit Price"))

print(MUnit)


#Mean

MePrice <- mean(sales$UnitPrice)
MeSales <- mean(sales$TotalSales)
MeQty <- mean(sales$Quantity)

sales$Year <- as.Date(sales$OrderDate, format = "%m/%d/%Y")
class(sales$Year)
sales$Year <- format(sales$Year, "%Y")

annualsales <- sales %>%
  group_by(Year) %>%
  summarise(mean(TotalSales))

annualprice <- sales %>%
  group_by(Year) %>%
  summarise(mean(UnitPrice))
            
annualqty <- sales %>%
  group_by(Year) %>%
  summarise(mean(Quantity))


barplot(height = annualsales$`mean(TotalSales)`, names = annualsales$Year,
        xlab = "Year", 
        ylab = 'Sales',
     main = paste("Mean of Sales X Year"))

barplot(height = annualprice$`mean(UnitPrice)`, names = annualsales$Year,
        xlab = "Year", 
        ylab = 'Sales',
     main = paste("Mean of Unit Price X Year"))

barplot(height = annualqty$`mean(Quantity)`, names = annualsales$Year,
        xlab = "Year", 
        ylab = 'Sales',
     main = paste("Mean of Quantity X Year"))



#Median 

Qty <- boxplot(sales$Quantity, 
              main = paste("Median of Quantity")
              )
abline (h = mean(sales$Quantity), col='red', lwd =2)
print(mean(sales$Quantity))

Qty <- boxplot(sales$UnitPrice, 
              main = paste("Median of Price")
              )
abline (h = mean(sales$UnitPrice), col= 'red', lwd =2)
print(mean(sales$UnitPrice))

Qty <- boxplot(sales$TotalSales, 
              main = paste("Median of Total Sales")
              )
abline (h = mean(sales$TotalSales), lwd =2)
print(median(sales$TotalSales))
print(median(sales$UnitPrice))
print(median(sales$Quantity))

```
Show the measures of dispersion, visually and numerically (where appropriate). Interpret those values in Markdown.
I used standard deviation to show the measure of dispersion by Region 
```{r}

sales$Year.factor <- factor(sales$Year)
sales$region.factor <- factor(sales$Region)
sales$prod.factor <- factor(sales$Product)

x <- sales %>%
  group_by(region.factor) %>%
  summarise(sd(Quantity))
plot(x, sales$region.factor, 
     xlab = 'Region',
     ylab = 'Standard Deviation')
print(sd(sales$Quantity))

y<- sales %>%
  group_by(region.factor) %>%
  summarise(sd(UnitPrice))
plot(y, sales$region.factor, 
     xlab = 'Region',
     ylab = 'Standard Deviation')
print(sd(sales$UnitPrice))

z <- sales %>%
  group_by(region.factor) %>%
  summarise(sd(TotalSales))
plot(z, sales$region.factor, 
     xlab = 'Region',
     ylab = 'Standard Deviation')
print(var(sales$TotalSales))

```
Visually depict the Interquartile range. Interpret the IQR for the dataset (in Markdown).
I used the IQR to understand the IQR for the Total Sales, Unit Price, and Quantity. At first, I tried region but by Product made more sense to provide as the IQR. 
```{r}

sales$region.factor <- factor(sales$Region)
sales$prod.factor <- factor(sales$Product)

#TotalSales 
IQR_Region <- aggregate(sales$TotalSales, by=list(Regs = sales$region.factor), FUN = IQR)
print(IQR_Region)

boxplot(x~ Regs, data = IQR_Region,
        main = "Total Sales IQR by Region",
        xlab = 'Region',
        ylab = 'IQR Total Sales')
abline(h = mean(IQR_Region$x), col = 'red', lwd = 3)
barplot(height = IQR_Region$x, names = IQR_Region$Regs)

#UnitPrice
IQR_Region1 <- aggregate(sales$UnitPrice, by=list(Regs = sales$prod.factor), FUN = IQR)
print(IQR_Region1)

boxplot(x~ Regs, data = IQR_Region1,
        main = "Unit Price IQR by Region",
        xlab = 'Products',
        ylab = 'IQR Unit Price')
abline(h = mean(IQR_Region1$x), col = 'red', lwd = 3)

barplot(height = IQR_Region1$x, names = IQR_Region1$Regs)
#Quantity 

IQR_Region2 <- aggregate(sales$Quantity, by=list(Regs = sales$prod.factor), FUN = IQR)
print(IQR_Region2)

boxplot(x~ Regs, data = IQR_Region2,
        main = "Quantity IQR by Region",
        xlab = 'Products',
        ylab = 'IQR Quantity')
abline(h = mean(IQR_Region2$x), col = 'red', lwd = 3)

barplot(height = IQR_Region2$x, names = IQR_Region2$Regs)
```
Create the statistics moments for the dataset.
Created the stats moment for Quantity, Unit Price, and Total Sales.
```{r}
# Youtube : https://youtu.be/gT5MXDzrN88?feature=shared

#Quantity
mean(sales$Quantity)
median(sales$Quantity)
sd(sales$Quantity)


ggplot(sales, aes(x = Quantity)) +
  ggtitle('Stats Moments') +
  labs(subtitle = 'Stats Moments of Quantity') +
  geom_histogram(fill = 'gray') +
  geom_vline(linetype = 'dashed', color = 'blue', xintercept = mean(sales$Quantity)) +
  geom_vline(linetype = 'dashed', color = 'orange', xintercept = median(sales$Quantity)) +
  geom_vline(linetype = 'dashed', color = 'green', xintercept = sd(sales$Quantity))


#UnitPrice

mean(sales$UnitPrice)
median(sales$UnitPrice)
sd(sales$UnitPrice)

ggplot(sales, aes(x = UnitPrice)) +
  ggtitle('Stats Moments') +
  labs(subtitle = 'Stats Moments of Unit Price') +
  geom_histogram(fill = 'gray') +
  geom_vline(linetype = 'dashed', color = 'blue', xintercept = mean(sales$UnitPrice)) +
  geom_vline(linetype = 'dashed', color = 'orange', xintercept = median(sales$UnitPrice)) +
  geom_vline(linetype = 'dashed', color = 'green', xintercept = sd(sales$UnitPrice))
 

#TotalSales

mean(sales$TotalSales)
median(sales$TotalSales)
sd(sales$TotalSales)

ggplot(sales, aes(x = TotalSales)) +
  ggtitle('Stats Moments') +
  labs(subtitle = 'Stats Moments of Total Sales') +
  geom_histogram(fill = 'gray') +
  geom_vline(linetype = 'dashed', color = 'blue', xintercept = mean(sales$TotalSales)) +
  geom_vline(linetype = 'dashed', color = 'orange', xintercept = median(sales$TotalSales)) +
  geom_vline(linetype = 'dashed', color = 'green', xintercept = sd(sales$TotalSales))
  labs(caption = "blue line is mean and orange is median.")

```

Experiment with histogram binning for the categorical data in R. This is open-ended intentionally because this dataset isn’t ideal for binning, but I want you to build something using binning just to see how it works.

```{r}
# sales$region.num <- factor(sales$Region)
# sales$prod.num <- factor(sales$Product)
# sales$Date.num <- factor(sales$OrderDate)
# sales$ID.num <- factor(sales$OrderID)

# https://stackoverflow.com/questions/34867308/how-to-make-grouped-bins-in-histogram-in-r-with-a-variable-other-than-frequency
cut_number(sales$TotalSales, 
           n = 10) %>%
          table()
sales <- sales %>%
  mutate(SalesCut = cut_number(sales$TotalSales, 
                              n = 50))


salesbin <- sales %>%
  group_by(SalesCut, TotalSales, Product, UnitPrice, Quantity) %>%
  summarise(avgTotal = mean(sales$TotalSales))


ggplot(salesbin, aes(x = Product,
                  y = TotalSales,
                  group=salesbin$SalesCut,
                  fill = factor(salesbin$SalesCut))) +
    geom_bar(stat = 'identity')

ggplot(salesbin, aes(x = Product,
                  y = UnitPrice,
                  group=salesbin$SalesCut,
                  fill = factor(salesbin$SalesCut))) +
  geom_bar(stat = 'identity')

```




